### 2020-05-15 Organismal_paper/Organismal_manuscript_and_code.Rmd
cleanup writing, add phenotypic integration. Make supplementary materials RMD

```{r setup, include=FALSE, echo=FALSE, eval=FALSE}
	# Load packages
	## General
library(knitr)
library(tidyverse)
library(stringr)
library(reshape2)
	## Biological
library(Rphylip)
library(arbutus)
library(vegan)
library(ape)
library(phangorn)
library(phytools)
library(picante)
library(geiger)
library(phylobase)
library(fields)
library(phylosignal)
library(geomorph)
library(phylopath)
library(phylolm)
library(convevol)
	## Graphics
library(FactoMineR)
library(factoextra)
library(corrplot)
library(BAMMtools)
library(gridExtra)
library(xtable)
library(colorRamps)
library(ggfortify)
library(ggpubr)

@@

Siphonophore tentilla (tentacle side branches) are unique biological structures for prey capture, composed of a complex arrangement of cnidocytes (stinging cells) bearing different types of nematocysts (stinging capsules) and auxiliary structures. Tentilla present an extreme morphological diversity of form and function across species. While associations between tentilla form and diet have been reported, the evolutionary history giving rise to this morphological diversity is largely unexplored. Here we explore the evolutionary gains and losses of novel tentilla substructures on the most recent siphonophore phylogeny. Tentilla have a precisely coordinated high-speed strike mechanism of synchronous unwinding and nematocysts discharge. Here we characterize the kinematic diversity of this prey capture reaction using high-speed video and find relationships with morphological characters. Since tentillum discharge occurs in synchrony across a broad morphological diversity, we evaluate how phenotypic integration is maintaining character correlations across evolutionary time. Moreover, we analyze the dimensionality of the tentillum morphospace, identify instances of heterochrony and morphological convergence, and generate hypotheses on the diets of understudied siphonophore species. Our findings indicate that siphonophore tentilla are phenotypically integrated structures with a complex evolutionary history leading to a phylogenetically structured diversity of forms which are predictive of kinematic performance and feeding habits.

@@

Siphonophores have fascinated zoologists for centuries for their extremely subspecialized colonial organization and integration. However, far less attention has been devoted to their unique tentacular structures: the tentilla. Like many cnidarians, siphonophore tentacles bear side branches (tentilla) with nematocysts. But unlike other cnidarians, most siphonophore tentilla are dynamic structures that react to prey encounters by shooting the nematocyst battery to slap around the prey. This maximizes the surface area of contact between the nematocysts and the prey they fire upon. Siphonophore tentilla present an exuberant diversity of morphologies, sizes, and nematocyst complements (Figure 1).

@@

In [@damian2019evolution], we collected the most extensive morphological dataset on siphonophore tentilla and nematocysts using state-of-the-art microscopy techniques, and expanded the taxon sampling of the phylogeny to disentangle the evolutionary history. The analyses we carried out led to novel insights into the evolution of predatory specialization. However, this work left behind many fascinating discoveries on the evolutionary history of tentilla morphology, including nematocyst novelties, phenotypic integration, convergent evolution, paedomorphosis, and predictions on diet using morphology. Here, we present these findings and shed light on the natural history of siphonophores.

@@

We explore the evolutionary history and diversification of tentilla morphologies using phylogenetic comparative methods on the morphological dataset and phylogeny published in [@damian2019evolution].  We fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. These models include a non-phylogenetic whie-noise model (WN), a neutral divergence Brownian Motion model (BM), an early-burst decreasing rate model (EB), and an Ornstein-Uhlenbeck model with stabilizing selection around a fitted optimum trait value. We then ranked the models in order of increasing parametric complexity (WN, BM, EB, OU), and compared the corrected Akaike Information Criterion (AICc) support scores [@sugiura1978further] to the lowest (best) score, using a cutoff of 2 units to determine significantly better support. When the best fitting model was not significantly better than a less complex alternative, we selected the least complex model (SM13). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (SM14). We calculated phylogenetic signals in each of the measured characters using Blomberg’s K [@blomberg2003testing] (SM13).

In order to explore the correlational structure among continuous characters and among their evolutionary histories, we used principal component analysis (PCA) and phylogenetic PCA [@revell2012phytools]. Since the character dataset contains many gaps due to missing characters and inapplicable states, we carried out these analyses on a subset of species and characters that allowed for the most complete dataset. This was done by removing the terminal filament characters (which are only shared by a small subset of species), and then removing species which had inapplicable states for the remaining characters. In addition, we obtained the correlations between the phylogenetic independent contrasts [@felsenstein1985phylogenies] using the package rphylip [@revell2014rphylip].

@@

![\label{figure3} Heatmap summarizing the morphological diversity measured in Damian-Serrano *et al.* 2020 for 96 species of siphonophores clustered by similarity (raw data in Dryad repository). Missing values from absent characters presented as dark grey cells, missing values produced from technical difficulties presented as white cells. Values scaled by character.](~/tentilla_morph/Figures_final/figure3.jpg)

*Evolutionary history of tentillum morphology* – [@damian2019evolution] produced the most speciose siphonophore molecular phylogeny to date, while incorporating the most recent findings in siphonophore deep node relationships. This phylogeny revealed for the first time that the genus *Erenna* is the sister to *Stephanomia amphytridis*. *Erenna* and *Stephanomia* bear the largest tentilla among all siphonophores, thus their monophyly indicates that there was a single evolutionary transition to giant tentilla. Siphonophore tentilla range in size from ~30 µm in some *Cordagalma* specimens to 2-4 cm in *Erenna* species, and up to 8 cm in *Stephanomia amphytridis* [@pugh2014review]. Most siphonophore tentilla measure between 175 and 1007 µm (1st and 3rd quartiles), with a median of 373 µm. The extreme gain of tentillum size in this newly found clade may have important implications for access to large prey size classes.

  Siphonophore tentilla are defined as lateral, monostichous evaginations of the tentacle gastrovascular lumen with epidermal nematocysts [@totton1965synopsis]. The buttons on *Physalia* tentacles were not traditionally regarded as tentilla, but [@bardi2007taxonomic] and our observations [@munro2018improved], confirm that the buttons contain evaginations of the gastrovascular lumen, thus satisfying all the criteria for the definition. In this light, and given that most Cystonectae bear conspicuous tentilla, we conclude (in agreement with [@munro2018improved]) that tentilla are likely ancestral to all siphonophores, and secondarily lost in *Apolemia* and *Bathyphysa conifera*.

In order to gain a broad perspective on the evolutionary history of tentilla, we reconstructed the phylogenetic positions of the main categorical character shifts using stochastic character mapping (SM22-30) and summarized in \@ref(figure7). Some of these characters include the gain and loss of nematocyst types. Based on external information, we assume that haploneme nematocysts are ancestrally present in siphonophore tentacles since they are present in the tentacles of many other hydrozoans. Haplonemes first diverged into spherical isorhizas of 2 size classes in Cystonectae, and elongated anisorhizas of one size class in Codonophora. Haplonemes were likely lost in the tentacles of *Apolemia* but retained as spherical isorhizas in other *Apolemia* tissues [@siebert2013re]. Similarly, while heteronemes exist in other tissues of cystonects, they appear in the tentacles of codonophorans exclusively, as birhopaloids in *Apolemia*, stenoteles in eucladophoran physonects, and microbasic mastigophores in calycophorans.

  The clades defined in [@damian2019evolution] are characterized by unique evolutionary innovations in their tentilla. The clade Eucladophora (containing Pyrostephidae, Euphysonectae, and Calycophorae) encompasses all of the extant Siphonophore species (178 of 186) except Cystonects and *Apolemia*. Innovations that arose along the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type. These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (Fig. \@ref(figure2)). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have evolved a wide variation in nematocyst and cnidoband sizes.

![\label{figure7} Siphonophore cladogram with the main categorical character gains (green) and losses (red) mapped. Some branch lengths were modified from the Bayesian chronogram to improve readability. The main visually distinguishable tentillum types are sketched next to the species that bear them, showing the location and arrangement of the main characters. In large, complex-shaped euphysonect tentilla, haplonemes were omitted for simplification. The rhizophysid *Bathyphysa conifera* branch was appended manually as a polytomy (dashed line).](~/tentilla_morph/Figures_final/figure7.jpg)

*Evolution of tentillum and nematocyst characters* – One-third of the characters measured in [@damian2019evolution] support a non-phylogenetic generative model, indicating they are not phylogenetically conserved (SM\@ref(ModelSupport)). Most (74%) characters present a significant phylogenetic signal, yet only total nematocyst volume, haploneme length, and heteroneme-to-cnidoband length ratio had a phylogenetic signal with K larger than 1. Total nematocyst volume and cnidoband-to-heteroneme length ratio showed strongly conserved phylogenetic signals. The majority (67%) of characters support BM models, indicating a history of neutral constant divergence. No relationship between phylogenetic signal and BM model support was found. Haploneme nematocyst length is the only character with support for an EB model of decreasing rate of evolution with time. No character had support for a single-optimum OU model (when uninformed by feeding guild regime priors).

![\label{figure8} A. Correlogram showing strength of ordinary (upper triangle) and phylogenetic (lower triangle) correlations between characters. Both size and color of the circles indicate the strength of the correlation (R^2^). B. Scatterplot of phylogenetic correlation against ordinary correlation showing a strong linear relationship (R^2^ = 0.92, 95% confidence between 0.90 and 0.93). Light red and blue boxes indicate congruent negative and positive correlations respectively. Darker red and blue boxes indicate strong (<-0.5 or >0.5) negative and positive correlation coefficients respectively.](~/tentilla_morph/Figures_final/figure8.jpg){height=600px}
*Evolution of nematocyst shape* – The greatest evolutionary change in haploneme nematocyst shape occurred in a single shift towards elongation in the stem of Tendiculophora, which contains the majority of described siphonophore species other than Cystonects, *Apolemia*, and Pyrostephidae. There is one secondary return to more oval, less elongated haplonemes in *Erenna*, but it does not reach the sphericity present in Cystonectae or Pyrostephidae (Fig. \@ref(figure10)). Heteroneme evolution presents a less discrete evolutionary history, where Tendiculophora evolved more elongate heteronemes, but the difference between theirs and other siphonophores is much smaller than the variation in shape within Tendiculophora, bearing no phylogenetic signal. In this clade, the evolution of heteroneme shape has diverged in both directions, and there is no correlation with haploneme shape (Fig. \@ref(figure10)), which has remained fairly constant (elongation between 1.5 and 2.5).

![\label{figure10} Phylomorphospace showing haploneme and heteroneme elongation (log scaled). Orange area delimits rod-shaped haplonemes, the blue area covers oval and round-shaped haplonemes. Smaller dots and lines represent phylogenetic relationships and ancestral states of internal nodes under BM. Species nodes in red lack either haplonemes or heteronemes, and their values are projected onto the axis of the nematocyst type they bear. Cystonects have no tentacle heteronemes and are projected onto the haploneme axis. Apolemiids have no tentacle haplonemes and are projected onto the heteroneme axis. ](~/tentilla_morph/Figures_final/figure10.jpg)

*Phenotypic integration of the tentillum* – Of the phylogenetic correlations (SM\@ref(figure8)a, lower triangle), 81.3% were positive and 18.7% were negative, while of the ordinary correlations (SM\@ref(figure8)a, upper triangle) 74.6% were positive and 25.4% were negative. Half (49.9%) of phylogenetic correlations were >0.5, while only 3.6% are < -0.5. Similarly, of the across-species correlations, 49.1% were >0.5 and only 1.5% were < -0.5. We found that 13.9% of character pairs had opposing phylogenetic and ordinary correlation coefficients. Just 4% have negative phylogenetic and positive ordinary correlations (such as rhopaloneme elongation ~ heteroneme-to-cnidoband length ratio and haploneme elongation, or haploneme elongation ~ heteroneme number), and only 9.9% of character pairs had positive phylogenetic correlation yet negative ordinary correlation (such as heteroneme elongation ~ cnidoband convolution and involucrum length, or rhopaloneme elongation with cnidoband length). These disparities can be caused by Simpson’s paradox [@blyth1972simpson]: the reversal of the sign of a relationship when a third variable (or a phylogenetic topology [@uyeda2018rethinking]) is considered. However, no character pair had correlation coefficient differences larger than 0.64 between ordinary and phylogenetic correlations (heteroneme shaft extension ~ rhopaloneme elongation has a Pearson’s correlation of 0.10 and a phylogenetic correlation of  -0.54). Rhopaloneme elongation shows the most incongruencies between phylogenetic and ordinary correlations with other characters.

@@

  In the non-phylogenetic PCA morphospace using only simple characters (SM\@ref(figure9)), PC1 (aligned with tentillum and tentacle size) explained 69.3% of the variation in the tentillum morphospace, whereas PC2 (aligned with heteroneme length, heteroneme number, and haploneme arrangement) explained 13.5%. In a phylogenetic PCA, 63% of the evolutionary variation in the morphospace is explained by PC1 (aligned with shifts in tentillum size), while 18% is explained by PC2 (aligned with shifts in heteroneme number and involucrum length).

![\label{morphodiet}Phylomorphospace of continuous morphological characters with inapplicable states transformed to zeroes, overlapped with polygons conservatively defining the space occupied by each feeding guild. ](~/tentilla_morph/Organismal_paper/figures/morphospace_diet.png){height=600px}

>>>>

```{r setup, include=FALSE, echo=FALSE, eval=FALSE}
	# Load packages
	## General
library(knitr)
library(tidyverse)
library(stringr)
library(reshape2)
	## Biological
library(Rphylip)
library(arbutus)
library(vegan)
library(ape)
library(phangorn)
library(phytools)
library(picante)
library(geiger)
library(phylobase)
library(fields)
library(phylosignal)
library(geomorph)
library(phylopath)
library(phylolm)
library(convevol)
library(surface)
	## Graphics
library(FactoMineR)
library(factoextra)
library(corrplot)
library(BAMMtools)
library(gridExtra)
library(xtable)
library(colorRamps)
library(ggfortify)
library(ggpubr)

@@

Siphonophore tentilla (tentacle side branches) are unique biological structures for prey capture, composed of a complex arrangement of cnidocytes (stinging cells) bearing different types of nematocysts (stinging capsules) and auxiliary structures. Tentilla present an exuberant morphological diversity of form and function across species. While associations between tentilla form and diet have been reported, the evolutionary history giving rise to this morphological diversity is largely unexplored. Here we explore the evolutionary gains and losses of novel tentillum substructures and nematocyst types on the most recent siphonophore phylogeny. Tentilla have a precisely coordinated high-speed strike mechanism of synchronous unwinding and nematocysts discharge. Here we characterize the kinematic diversity of this prey capture reaction using high-speed video and find relationships with morphological characters. Since tentillum discharge occurs in synchrony across a broad morphological diversity, we evaluate how phenotypic integration is maintaining character correlations across evolutionary time. Moreover, we analyze the dimensionality of the tentillum morphospace, identify instances of heterochrony and morphological convergence, and generate hypotheses on the diets of understudied siphonophore species. Our findings indicate that siphonophore tentilla are phenotypically integrated structures with a complex evolutionary history leading to a phylogenetically structured diversity of forms which are predictive of kinematic performance and feeding habits.

@@

SSiphonophores have fascinated zoologists for centuries for their extremely subspecialized colonial organization and integration. Today we hold more knowledge than ever on the morphological diversity of this group due to the extensive work of siphonophore taxonomists in the past few decades (cite Phil, Steve, etc), which has been elegantly synthesized in detailed synopsis [@totton1954synopsis, @mapstone2014global]. In addition, recent advances in phylogenetic analyses of siphonophores [@munro2018improved;@damian2019evolution] have provided a macroevolutionary context to interpret this diversity. With these assets in hand, we can now begin to study siphonophores from an orthogonal perspective, focusing on the diversity and evolutionary history of specific structures. Here we focus on one of such structures: the tentilla. Like many cnidarians, siphonophore tentacles bear side branches (tentilla) with nematocysts. But unlike other cnidarians, most siphonophore tentilla are dynamic structures that react to prey encounters by shooting the nematocyst battery to slap around the prey. This maximizes the surface area of contact between the nematocysts and the prey they fire upon. In addition, siphonophore tentilla present a remarkable diversity of morphologies, sizes, and nematocyst complements (Fig, /@ref(figure2)). Our overarching aim is to organize all this phenotypic diversity in a phylogenetic context, and identify the evolutionary processes that generated it.

@@

In [@damian2019evolution], we collected the most extensive morphological dataset on siphonophore tentilla and nematocysts using state-of-the-art microscopy techniques, and expanded the taxon sampling of the phylogeny to disentangle the evolutionary history. The analyses we carried out led to novel, generalizable insights into the evolution of predatory specialization. The work we present here is complementary to [@damian2019evolution], showcasing a far more detailed account on the evolutionary history of tentilla morphology.

Nematocysts are unique biological weapons for defense and prey capture exclusive to the phylum Cnidaria. Mariscal ([@mariscal1974nematocyst]) reported that hydrozoans have the largest diversity of nematocyst types among cnidarians. Among them, siphonophores present the greatest variety of types [@mapstone2014global], and vary widely across taxa in which and how many types they carry on their tentacles. While recent descriptive studies have expanded and confirmed our understanding of this diversity, the evolutionary history of nematocyst type gain and loss in siphonophores remains unexplored. Thus, here we reconstruct the evolution of shifts, gains, and losses of nematocyst types, subtypes, and other major categorical traits that led to the extant diversity we see in siphonophore tentilla. In [@damian2019evolution], we found strong associations between piscivory and haploneme shape. These associations could have been produced by convergent changes in the adaptive optima of these characters. Here we set out to test this hypothesis using comparative model fitting methods. Analyzing the diversity of morphological states from a phylogenetic perspective allows us to identify the specific evolutionary processes that gave rise to it. Here we fit and compare a variety of macroevolutionary models to siphonophore tentilla morphology measurement data to identify instances of neutral divergence, stabilizing selection, changes in the speed of evolution, convergent evolution, and paedomorphosis (larval characters present in mature colonies).

In [@damian2019evolution] we fitted discriminant analyses to identify characters that are predictive of feeding guild. These discriminant analyses can be used to generate hypotheses on the diets of ecologically understudied siphonophore species for which we have morphology data. Here we present a Bayesian prediction for the feeding guild of 45 species using the morphological dataset in [@damian2019evolution]. As mentioned above, tentilla are far from being ornamental shapes and are in fact violently reactive weapons for prey capture [@Mackie:1987uy;@damian2019evolution]. While we now have detailed characterizations of tentilla morphologies across many species, the diversity of dynamic performances and their relationships to the undischarged morphologies have not been examined to date. To address this gap, we set out to record high-speed video of the *in vivo* discharge dynamics of several siphonophore species at sea, and compare the kinematic attributes to their morphological characters.

@@

All character data and the phylogeny analyzed here were published in [@damian2019evolution]. We log transformed all the continuous characters that did not pass Shapiro-Wilks normality tests, and used the ultrametric constrained Bayesian time tree in all comparative analyses. When missing data was incorporated as inapplicable states, we removed those species with characters that could not be measured due to technical limitations. We used the feeding guild categories detailed in [@damian2019evolution] with one modification: including all *Forskalia* spp. as generalists not only a single *Forskalia* species on the tree after a reinterpretation of the data in [@purcell1981dietary]. In order to characterize the evolutionary history of tentilla morphology, we fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. These models include a non-phylogenetic white-noise model (WN), a neutral divergence Brownian Motion model (BM), an early-burst decreasing rate model (EB), and an Ornstein-Uhlenbeck (OU) model with stabilizing selection around a fitted optimum trait value. In the same way as [@damian2019evolution] we then ordered the models by increasing parametric complexity (WN, BM, EB, OU), and compared their corrected Akaike Information Criterion (AICc) scores [@sugiura1978further]. We used the lowest (best) score using a delta cutoff of 2 units to determine significance relative to the next simplest model (SM10). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (SM11). We calculated phylogenetic signals in each of the measured characters using Blomberg’s K [@blomberg2003testing] (SM10). To reconstruct the ancestral character states of nematocyst types and other categorical traits, we use stochastic character mapping (SIMMAP) using the package *phytools* [@revell2012phytools].

In order to examine the phenotypic integration in the tentillum, we explored the correlational structure among continuous characters and among their evolutionary histories using principal component analysis (PCA) and phylogenetic PCA [@revell2012phytools]. Since the character dataset contains many gaps due to missing characters and inapplicable states, we carried out these analyses on a subset of species and characters that allowed for the most complete dataset. This was done by removing the terminal filament characters (which are only shared by a small subset of species), and then removing species which had inapplicable states for the remaining characters (apolemiids and cystonects). In addition, we obtained the correlations between the phylogenetic independent contrasts [@felsenstein1985phylogenies] using the package *rphylip* [@revell2014rphylip]. We identified four hypothetical modules among the tentillum characters: (1) The tentillum scaffold module -- cnidoband length & width, nematocyst row number, pedicle & elastic strand width, tentacle width; (2) the heteroneme module -- heteroneme length & width, shafts length & width; (3) the haploneme module -- length and width; and (4) the terminal filament module -- desmoneme & rhopaloneme length & width. To test an quantify phenotypic integration between these multivariate modules, we used the phylogenetic phenotypic integration test in the package *geomorph* [@adams2016geomorph].

When looking at the morphospace of species in different feeding guilds, we also used PCA on the complete tentacular character dataset transforming inapplicable states of absent characters to zeros to account for similarity based on character presence/absence. Using these principal components, we examined the occupation of the morphospace across species in different feeding guilds using a phylogenetic MANOVA with the package *geiger* [@harmon2008geiger] to assess the variation explained, and a morphological disparity test with the package *geomorph* [@adams2016geomorph] to assess differences in the extent occupied by each guild.

In order to detect and evaluate instances of convergent evolution, we used the package SURFACE [@ingram2013surface]. This tool identifies OU regimes and their optima given a tree and character data, and then evaluates where the same regime has appeared independently in different lineages. We applied these analyses to the haploneme nematocyst length and width characters as well as to the most complete dataset without inapplicable character states.

In order to generate hypotheses on the diets of siphonophores using tentilla morphology, we used the discriminant analyses of principal components (DAPC) trained in [@damian2019evolution] using the package *adegenet* [@jombart2008adegenet] to predict the feeding guilds of species in the dataset for which there are no published feeding observations.

In order to observe the discharge behavior of different tentilla, we recorded high speed footage (1000-3000 fps) of tentillum and nematocyst discharge by live siphonophore specimens (26 species) using a Phantom Miro 320S camera mounted on a stereoscopic microscope. We mechanically elicited tentillum and nematocyst discharge using a fine metallic pin. We used the Phantom PCC software to analyze the footage. For the 10 species recorded, we measured total cnidoband discharge time (ms), heteroneme filament length (μm), and discharge speeds (mm/s) for cnidoband, heteronemes, haplonemes, and heteroneme shafts when possible (data available in the Dryad repository).

@@

![\label{figure3} Heatmap summarizing the morphological diversity measured in Damian-Serrano *et al.* 2020 for 96 species of siphonophores clustered by similarity (raw data published in Damian-Serrano 2020). Missing values from absent characters presented as dark grey cells, missing values produced from technical difficulties presented as white cells. Values scaled by character. ](~/tentilla_morph/Figures_final/figure3.jpg)

*Evolutionary history of tentillum morphology* – In [@damian2019evolution], we produced the most speciose siphonophore molecular phylogeny to date, while incorporating the most recent findings in siphonophore deep node relationships. This phylogeny revealed for the first time that the genus *Erenna* is the sister to *Stephanomia amphytridis*. *Erenna* and *Stephanomia* bear the largest tentilla among all siphonophores, thus their monophyly indicates that there was a single evolutionary transition to giant tentilla. Siphonophore tentilla range in size from ~30 µm in some *Cordagalma* specimens to 2-4 cm in *Erenna* species, and up to 8 cm in *Stephanomia amphytridis* [@pugh2014review]. Most siphonophore tentilla measure between 175 and 1007 µm (1st and 3rd quartiles), with a median of 373 µm. The extreme gain of tentillum size in this newly found clade may have important implications for access to large prey size classes such as adult deep-sea fishes.

  Siphonophore tentilla are defined as lateral, monostichous evaginations of the tentacle gastrovascular lumen with epidermal nematocysts [@totton1965synopsis]. The buttons on *Physalia* tentacles were not traditionally regarded as tentilla, but [@bardi2007taxonomic] and our observations [@munro2018improved], confirm that the buttons contain evaginations of the gastrovascular lumen, thus satisfying all the criteria for the definition. In this light, and given that most Cystonectae bear conspicuous tentilla, we conclude (in agreement with [@munro2018improved] and [@damian2019evolution]) that tentilla are likely ancestral to all siphonophores, and secondarily lost twice, once in *Apolemia* and again in *Bathyphysa conifera*.

  In order to gain a broad perspective on the evolutionary history of tentilla, we reconstructed the phylogenetic positions of the main categorical character shifts using stochastic character mapping (SM1-9) and summarized in (Fig. \@ref(figure7)). Some of these characters include the gain and loss of nematocyst types. Based on external information, we assume that haploneme nematocysts are ancestrally present in siphonophore tentacles since they are present in the tentacles of many other hydrozoans [@mariscal1974nematocysts]. Haplonemes first diverged into spherical isorhizas of 2 size classes in Cystonectae, and elongated anisorhizas of one size class in Codonophora. Haplonemes were likely lost in the tentacles of *Apolemia* but retained as spherical isorhizas in other *Apolemia* tissues [@siebert2013re]. Similarly, while heteronemes exist in other tissues of cystonects, they appear in the tentacles of codonophorans exclusively, as birhopaloids in *Apolemia*, stenoteles in eucladophoran physonects, and microbasic mastigophores in calycophorans.

  The clades defined in [@damian2019evolution] are characterized by unique evolutionary innovations in their tentilla. The clade Eucladophora (containing Pyrostephidae, Euphysonectae, and Calycophorae) encompasses all of the extant Siphonophore species (178 of 186) except Cystonects and *Apolemia*. Innovations that arose along the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type. These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (Fig. \@ref(figure2)). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have evolved a wide variation in nematocyst and cnidoband sizes. Ancestrally (and retained in most prayomorphs and hippopodids), the calycophoran tentillum is recurved where the proximal and distal ends of the cnidoband are close together. Diphyomorph tentilla are slightly different in shape, with straighter cnidobands.

![\label{figure7} Siphonophore cladogram with the main categorical character gains (green) and losses (red) mapped. Some branch lengths were modified from the Bayesian chronogram to improve readability. The main visually distinguishable tentillum types are sketched next to the species that bear them, showing the location and arrangement of the main characters. In large, complex-shaped euphysonect tentilla, haplonemes were omitted for simplification. The hypothesized phylogenetic placement of the rhizophysid *Bathyphysa conifera* branch was appended manually as a polytomy (dashed line).](~/tentilla_morph/Figures_final/figure7.jpg)

*Evolution of tentillum and nematocyst characters* – One-third of the characters measured in [@damian2019evolution] support a non-phylogenetic generative model, indicating they are not phylogenetically conserved (SM10). Most (74%) characters present a significant phylogenetic signal, yet only total nematocyst volume, haploneme length, and heteroneme-to-cnidoband length ratio had a phylogenetic signal with K larger than 1. Total nematocyst volume and cnidoband-to-heteroneme length ratio showed strongly conserved phylogenetic signals. The majority (67%) of characters support BM models, indicating a history of neutral constant divergence. We did not find any relationship between phylogenetic signal and BM model support. Haploneme nematocyst length was the only character with support for an EB model of decreasing rate of evolution with time. No character had support for a single-optimum OU model (when notinformed by feeding guild regime priors).

![\label{figure8} A. Correlogram showing strength of ordinary (upper triangle) and phylogenetic (lower triangle) correlations between characters. Both size and color of the circles indicate the strength of the correlation (R^2^). B. Scatterplot of phylogenetic correlation against ordinary correlation showing a strong linear relationship (R^2^ = 0.92, 95% confidence between 0.90 and 0.93). Light red and blue boxes indicate congruent negative and positive correlations respectively. Darker red and blue boxes indicate strong (<-0.5 or >0.5) negative and positive correlation coefficients respectively.](~/tentilla_morph/Figures_final/figure8.jpg){height=600px}

*Evolution of nematocyst shape* – The greatest evolutionary change in haploneme nematocyst shape occurred in a single shift towards elongation in the stem of Tendiculophora, which contains the majority of described siphonophore species other than Cystonects, *Apolemia*, and Pyrostephidae. There is one secondary return to more oval, less elongated haplonemes in *Erenna*, but it does not reach the sphericity present in Cystonectae or Pyrostephidae (Fig. \@ref(figure10)). Heteroneme evolution presents a less discrete evolutionary history, where Tendiculophora evolved more elongate heteronemes, but the difference between theirs and other siphonophores is much smaller than the variation in shape within Tendiculophora, bearing no phylogenetic signal. In this clade, the evolution of heteroneme shape has diverged in both directions, and there is no correlation with haploneme shape (Fig. \@ref(figure10)), which has remained fairly constant (elongation between 1.5 and 2.5). Haploneme and heteroneme shape share 21% of their variance across extant values, and 53% of the variance in their shifts along the branches of the phylogeny. However, much of this correlation is due to the sharp contrast between Pyrostephidae and their sister group Tendiculophora. We searched for regime shifts in the evolution of haploneme nematocyst shape characters using a SURFACE [@ingram2013surface].  SURFACE identified seven distinct OU regimes in the evolutionary history of haploneme length and width (Fig. \@ref(surface)A). The different regimes are located (1) in cystonects, (2) in pyrostephids, (7) in apolemiids, (9) in *Erenna*, (6) in *Stephanomia*, (3) in most of Tendiculophora, (5) in *Cordagalma ordinatum*, (4) in most diphyomorphs, and (8) in *Abylopsis tetragona* and *Diphyes dispar*.

![\label{figure10} Phylomorphospace showing haploneme and heteroneme elongation (log scaled). Orange area delimits rod-shaped haplonemes, the blue area covers oval and round-shaped haplonemes. Smaller dots and lines represent phylogenetic relationships and ancestral states of internal nodes under BM. Species nodes in red lack either haplonemes or heteronemes, and their values are projected onto the axis of the nematocyst type they bear. Cystonects have no tentacle heteronemes and are projected onto the haploneme axis. Apolemiids have no tentacle haplonemes and are projected onto the heteroneme axis. ](~/tentilla_morph/Figures_final/figure10-01-01.jpg)

*Phenotypic integration of the tentillum* – Phenotypically integrated structures maintain evolutionary correlations between its constituent characters. Of the phylogenetic correlations (Fig. \@ref(figure8)a, lower triangle), 81.3% were positive and 18.7% were negative, while of the ordinary correlations (Fig. \@ref(figure8)a, upper triangle) 74.6% were positive and 25.4% were negative. Half (49.9%) of phylogenetic correlations were >0.5, while only 3.6% are < -0.5. Similarly, among the correlations across extant species, 49.1% were >0.5 and only 1.5% were < -0.5. In addition, we found that 13.9% of character pairs had opposing phylogenetic and ordinary correlation coefficients. Just 4% have negative phylogenetic and positive ordinary correlations (such as rhopaloneme elongation ~ heteroneme-to-cnidoband length ratio and haploneme elongation, or haploneme elongation ~ heteroneme number), and only 9.9% of character pairs had positive phylogenetic correlation yet negative ordinary correlation (such as heteroneme elongation ~ cnidoband convolution and involucrum length, or rhopaloneme elongation with cnidoband length). These disparities could be explained by Simpson’s paradox [@blyth1972simpson]: the reversal of the sign of a relationship when a third variable (or a phylogenetic topology [@uyeda2018rethinking]) is considered. However, no character pair had correlation coefficient differences larger than 0.64 between ordinary and phylogenetic correlations (heteroneme shaft extension ~ rhopaloneme elongation has a Pearson’s correlation of 0.10 and a phylogenetic correlation of  -0.54). Rhopaloneme elongation shows the most incongruencies between phylogenetic and ordinary correlations with other characters. The phenotypic integration test showed significant integration signal between all modules, tentillum and haploneme modules sharing the greatest regression coefficient (SM12).

@@

  In the non-phylogenetic PCA morphospace using only simple characters (Fig. \@ref(figure9)), PC1 (aligned with tentillum and tentacle size) explained 69.3% of the variation in the tentillum morphospace, whereas PC2 (aligned with heteroneme length, heteroneme number, and haploneme arrangement) explained 13.5%. In a phylogenetic PCA, 63% of the evolutionary variation in the morphospace is explained by PC1 (aligned with shifts in tentillum size), while 18% is explained by PC2 (aligned with shifts in heteroneme number and involucrum length).

![\label{morphodiet}Phylomorphospace showing PC1 and PC2 from a PCA of continuous morphological characters with inapplicable states transformed to zeroes, overlapped with polygons conservatively defining the space occupied by each feeding guild. Lines between species coordinates show the phylogenetic relationships between them. ](~/tentilla_morph/Organismal_paper/figures/morphospace_diets_full.png){height=600px}
